---
title: "GDP Colombia"
subtitle: "Annual and quarterly growth rates"
author: "Miguel Martinez"
date: today
format: 
  html:
    toc: true
    theme: cosmo
editor: visual
execute:
  warning: false
---

This document presents the process of visualizing the evolution of the annual and quarterly Colombian GDP. The data is made public by the Departamento Administrativo Nacional de Estadística ([DANE](https://www.dane.gov.co/)), and you can reproduce this exercise downloading the in the following links:

1.  [Quarterly data](https://www.dane.gov.co/index.php/estadisticas-por-tema/cuentas-nacionales/cuentas-nacionales-trimestrales/pib-informacion-tecnica)
2.  [Annual data](https://www.dane.gov.co/index.php/estadisticas-por-tema/cuentas-nacionales/cuentas-nacionales-anuales)

The first step is loading the required packages:

```{r}

library(readxl)
library(tidyverse)

```

## Load data

```{r}

# Original quarterly data

df1 = readxl::read_excel("anex-ProduccionConstantes-Itrim2025.xlsx",
                         sheet = "Cuadro 1",
                         range = "C45:CF62",
                         col_names = FALSE)

# Annual data

df2 = readxl::read_excel("anex-CuentasNalANuales-AgreMacroeconomicos-2023p.xlsx",
                         sheet = "Cuadro 2",
                         range = "E165:X269",
                         col_names = FALSE)

```

## Indexing

```{r}

# Quarterly

df1[1, 1] = "Año"
df1[2, 1] = "Trimestre"

# Annual

df2[1, 1] = "Año"
df2[1, 20] = "2023"
```

## Dtype transformation

```{r}

# We transform all columns to character in df2

df2 = df2 |> 
  dplyr::mutate_if(is.numeric, as.character)
```

## **Lengthening Data**

```{r}

# Quarterly

df1 = df1 |>
  tidyr::pivot_longer(cols = 6:82) |>
  tidyr::fill(value) |>
  tidyr::pivot_wider(names_from = 1,
                     values_from = value) |>
  dplyr::select(Año:`Producto Interno Bruto`)

```

```{r}

# Annual

df2 = df2 |>
  tidyr::pivot_longer(cols = 3:20) |>
  tidyr::pivot_wider(names_from = 1,
                     values_from = value,
                     values_fn = list) |>
  dplyr::select(Año, `Producto interno bruto`)

```

## D type transformation

```{r}

# We transform the variables to numeric in the quarterly data

df1 = df1 |>
  dplyr::mutate_at(vars(4:18), as.numeric)

```

```{r}

# Please notice that some years have strings, so we need to replace them first

años = df1 |>
  dplyr::distinct(Año)

años$Año

```

```{r}

# Now we transform the "string" format of years to "numeric"

df1 = df1 |>
  dplyr::mutate(
    Año =  stringr::str_remove_all(Año, "[pr]"),
    Año = as.numeric(Año))

glimpse(df1)

```

```{r}

# We transform all columns to numeric in df2

df2 = df2 |>
  dplyr::mutate_at(vars(1:2), as.numeric)

df2
```

## Filtering

```{r}

# We filter quarterly data from 2020

df1 = df1 |>
  dplyr::filter(Año >= 2020)

```

## Quarterly

```{r}

dashline = 0   # Baseline

df1 |> 
  ggplot(aes(x = Trimestre, y = `Producto Interno Bruto`, group = 1)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = dashline, color = "black", linewidth = 0.5, linetype = 2) +
  facet_grid( . ~ Año) +
  theme_classic() +
  labs(title = "Producto Interno Bruto (PIB)",
       subtitle = "Tasas de crecimiento anual en volumen",
       x = "Trimestres",
       y = "Variación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, PIB trimestral (coyuntural). A precios constantes") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0))

```

## Annually

```{r}

dashline = 0       # Baseline
shadedarea = 2020  # Year of interest

df2 |> 
  ggplot(aes(x = Año, y = `Producto interno bruto`)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = dashline, color = "black", linewidth = 0.5, linetype = 2) +
  geom_vline(xintercept = shadedarea, col = "grey", linewidth = 22, alpha = 0.4) +
  scale_x_continuous(breaks = seq(2006, 2024, by = 2)) +
  theme_bw() +
  labs(title = "Producto Interno Bruto (PIB)",
       subtitle = "(variación anual)",
       x = "Año",
       y = "Variación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, tasa de crecimiento en volumen") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0))

```

## Conclusion

We can see that the Colombian GDP is exposing signals of recovering from the pandemic shock. In the last quarters the growth is significant and positive but for now not enough for the expectation of 3% or higher displayed a decade before.
